---
title: "DA5020.P1"
authors: "Aditya Elayavalli, Mason Lewis, Nitin Dinesh"
date: "2024-02-10"
output: 
  html_document:
    theme: journal
    toc: yes
    toc_float: 
      collapsed: true
subparagraph: true
---

```{r message=FALSE}
library(tidyverse)
library(stringr)
library(ggplot2)
library(dplyr)
```

## Part 1

### Question 1
The below code will create the data frame. We will be putting it into the variable called doctor_data. 

```{r}
doctor_data <- data.frame(
  doctor_type = c("PCP", "Psychiatrist", "Surgeon", "Anesthesia"),
  doctor_lastname = c("Smith", "Dame", "Jones", "Zayas"),
  location = c("MA", "ME", "NH", "VT"),
  AVG_Rating = c(7, 9, 8, 9)
)
```

The below command will print the data frame we created above.

```{r}
print(doctor_data)
```


### Question 2

Can index to get the what is in row 1 and column 2.

```{r}
#selecting required row and column
doctor_data[1,2]
```

Similarly we can index to see what is in rows 2-4.

```{r}
doctor_data[2:4,]
```


The same method can be used to determine what is in the 4th column.

```{r}
doctor_data[,4]
```

### Question 3

To load the data set we can use data() function.

```{r}
data(mtcars)
```

The below function will let us see the top of the data set to ensure its been loaded properly.

```{r}
head(mtcars)
```

The below code chunk will provide the structure of the dataset. 

```{r}
glimpse(mtcars)
```

We are going to use the ggplot command to plot the scatterplot between the variables wt and mpg. 

```{r}
ggplot(mtcars,aes(x=wt, y=mpg)) +
  geom_point(color = "#1F4E78") +
  labs(title = "Scatter Plot depicting relationship of wt and mpg",
       x = "Weight (lbs)",
       y = "MPG",
       caption = "See how weight impacts MPG") +
  theme(text = element_text(family = "serif"),
          panel.grid.major.x = element_line(linewidth=0.1),
          panel.grid.major.y = element_line(linewidth = 0.3),
          plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          plot.caption = element_text(size = 6),
          plot.margin = margin(10, 20, 10, 20, "pt"),
          plot.background = element_rect(linewidth = 1, color = "black"))
```

Negative trend. We chose these variables as we wanted to see how weight effects the miles per gallon efficiency. As we can see there is a negative trend indicating heavier the car, lower the efficiency which makes sense.


### Question 4

The below command will tell us the min value, 1st quartile, median, mean, 3rd quartile and Max values in each column. 

```{r}
summary(mtcars)
```

The cor.test() can help calculate the pearson coefficient. A pearson coefficient is the most effective way to determine a linear correlation. The coefficient lies within the range of -1 and 1. Depending on where the coefficient lands in the range, one can determine the strength and direction of the relationship between two variables. The below code chunk will allow us to determine the coefficient. 

```{r}
cor.test(mtcars$wt, mtcars$mpg, 
                    method = "pearson")

```

As we can see the coefficient is -0.8676594. We can see that the variables therefore have a strong negative linear relationship, meaning as one variable increases the other decreases. This can also be seen with how scatterplot was produced. 



## Part 2

### Question 1

```{r message = FALSE}
data_url <- "https://data.ny.gov/api/views/ngbt-9rwf/rows.csv"
addiction_df <- read_csv(data_url)

# Confirm that the data has been properly imported 
```


### Question 2

```{r}
# Remove any rows which have NAs in any columns
addiction_df <- addiction_df %>%
  filter(complete.cases(.))

# Renaming column names to include underscores instead of white spaces
colnames(addiction_df) <- c("Year", "County_of_Program_Location", "Program_Category",
                            "Service_Type", "Age_Group", "Primary_Substance_Group", 
                            "Admissions")

# Creating a data frame which calculates the frequency of each instance of the different variables 
frequency_df <- addiction_df %>%
  
  # Use pivot_longer() to tidy the data into one "Variable Column"
  # Used values_transform to change the numeric columns into characters for compatibility
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value",
               values_transform = list(Value = as.character)) %>%
  
  # Group by Variable and Value
  # Summarize() to create a column with the frequency of each instance of each variable 
  group_by(Variable, Value) %>%
  summarize(Frequency = n())

# Check statements to ensure the data frame contains the desired data
head(frequency_df, 3)
dim(frequency_df)

# Create a box plot of this new data frame which will display outliers of each variable 
ggplot(frequency_df, aes(x = Variable, y = Frequency)) + 
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Frequency of Values of Variables in the Dataset",
       x = "Variable", y= "Frequency of Values of the Variables",
       caption = "Expose outliers that may exist within variables") +
  theme(text = element_text(family = "serif"),
        legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linewidth = 0.2),
        panel.grid.minor.y = element_blank(),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(size = 6),
        plot.margin = margin(10, 20, 10, 20, "pt"),
        plot.background = element_rect(linewidth = 1, color = "black"))

# Creating a frequency data frame to examine a variable of interest : Age_Group
age_count <- addiction_df %>%
 count(Age_Group) %>%
 arrange(desc(n))

age_count

# Creating a frequency data frame to examine a variable of interest : Primary_Substance_Group
substance_count <- addiction_df %>%
 count(Primary_Substance_Group) %>%
 arrange(desc(n))

substance_count

# Remove data where substance_group = 1
addiction_df <- addiction_df %>%
  filter(Primary_Substance_Group != "None")

# Creating a frequency data frame to examine a variable of interest : Service_Type
service_count <- addiction_df %>%
 count(Service_Type) %>%
 arrange(desc(n))

service_count
```


With admissions being the only true numeric column of the data, it is worth observing if there are any outliers which need to be removed from the data. An outlier has been designated as a value which is greater than three standard deviations from the mean. With this definition, any single observation with 411 admissions or greater is considered an outlier.

That being said, further investigation is needed to determine if these outliers should be considered invalid data. New York is a unique state, with sparsely populated rural areas and extreme population density in New York City. As a result, the county in which the outlier data is from is import context in determining if these observations should be used in further analysis. The five greatest outliers with the highest number of admissions not only come from New York County (Manhattan), but also are for treatment of alcohol dependency, the only federally legal substance included in the substance groups. The context surrounding the outliers suggests that rather than the extreme admissions values being a mistake, they're reflective of the population differences among counties in the state. This would be a different story if we observed that same value coming from Yates County, with a population of approximately 25,000.


```{r}
# Calculate mean and standard deviation of Admissions data
mean_admissions <- mean(addiction_df$Admissions)
sd_admissions <- sd(addiction_df$Admissions)

outlier_admissions <- addiction_df %>%
  mutate(z_score = abs(mean_admissions - Admissions) / sd_admissions) %>%

  # Filter out z-scores less than 3
  filter(z_score >= 3) %>%
  
  # Select and arrange Admissions to display
  select(Admissions) %>%
  arrange(desc(Admissions))

# Display biggest outliers
head(outlier_admissions, 5)

# Investigate and display biggest outliers
highest_admissions_observations <- addiction_df %>%
  arrange(desc(Admissions)) %>%
  slice_head(n = 5)

highest_admissions_observations
```

To further confirm this point, summary statistics have been presented. This low median value relative to the mean, signals the presence of extreme values to the high side, which we already know to be true. Additionally, the low median value reinforces the idea there are a large number of observations from smaller counties. The fact that the data is split across so many variables also makes it difficult to determine observations worthy of removal. To better clarify which admissions values may be truly extreme, the data would have to be separated so that only similar groups are being compared to one another. For example, it would only make sense to compare admission numbers of high-population counties for the same substance and age group. Alternatively, the addition of population data to the data frame could allow for per capita calculations, making comparison across groups more suitable.

```{r}
# Display summary statistics
mean_admissions
median(addiction_df$Admissions)
sd_admissions
```


### Question 3

```{r}
# Created a data frame which lists every county in alphabetical order
county_count <- addiction_df %>%
 count(County_of_Program_Location) %>%
 arrange(County_of_Program_Location)

# Assigned alphabetical list of counties to a vector: counties
counties <- county_count$County_of_Program_Location

# Hard coded an alphabetical list of county codes into a vector: codes
codes <- c('AL', 'AG', 'BX', 'BM', 'CA', 'CY', 'CH', 'CM', 'CN', 'CL',
           'CO', 'CR', 'DE', 'DU', 'ER', 'ES', 'FR', 'FU', 'GE', 'GR', 'HE', 'JE',
           'KN', 'LE', 'LI', 'MA', 'MO', 'MG', 'NA', 'NY', 'NI', 'ON', 'OD', 'OT',
           'OR', 'OL', 'OS', 'OG', 'PU', 'QU', 'RE', 'RM', 'RO', 'SL', 'SA', 'SC',
           'SH', 'SY', 'SE', 'ST', 'SU', 'SV', 'TI', 'TO', 'UL', 'WR', 'WS', 'WA',
           'WE', 'WY', 'YA')

# Created a data frame of county names and their respective county codes
county <- data.frame(County_of_Program_Location = counties, county_code = codes)

# Check that the data frame contains the necessary data
head(county, 3)

# Created a data frame which lists every program category in alphabetical order
program_count <- addiction_df %>%
 count(Program_Category) %>%
 arrange(Program_Category) 

# Assigned alphabetical list of programs to a vector: programs
programs <- program_count$Program_Category

# Used the substr() function to extract the first 3 characters in Program_Category for use as a code
# Created a data frame with programs and program codes
program_category <- data.frame(Program_Category = programs) %>%
  mutate(program_code = substr(Program_Category, 1, 3))

# Check Statement
head(program_category, 3)

# Alphabetical Substance List
substance_count <- addiction_df %>%
 count(Primary_Substance_Group) %>%
 arrange(Primary_Substance_Group) 

# Assignment to vector
substances <- substance_count$Primary_Substance_Group

# Extraction of first three characters as a code, creation of data frame
primary_substance_group <- data.frame(Primary_Substance_Group = substances) %>%
  mutate(substance_code = substr(Primary_Substance_Group, 1, 3))

# Check Statement
primary_substance_group

# Make new dataframe with age codes
age_groups <- c('Under 18', '18 through 24', '25 through 34', '35 through 44',
                '45 through 54', '55 and Older')

age_codes <- c('<18', '18-24', '25-34', '35-44', '45-54', '55+')

age_df <- data.frame(Age_Group = age_groups, age_code = age_codes)

# Created a copy of the original data frame
admissions_data <- data.frame(addiction_df)

# Merge data frames to replace full county names with county codes
admissions_data <- admissions_data %>%
  
  # Add column with county codes
  left_join(county, by = ("County_of_Program_Location")) %>%
  
  # Remove full names, rename county codes
  select(-County_of_Program_Location) %>%
  rename(County_of_Program_Location = county_code)

# Merge data frames to replace program names with program codes
admissions_data <- admissions_data %>%
  
  # Add column with county codes
  left_join(program_category, by = ("Program_Category")) %>%
  
  # Remove full names, rename program codes
  select(-Program_Category) %>%
  rename(Program_Category = program_code)

# Merge data frames to replace substance groups with substance codes
admissions_data <- admissions_data %>%
  
  # Add column with county codes
  left_join(primary_substance_group, by = ("Primary_Substance_Group")) %>%
  
  # Remove full names, rename substance codes, move columns to original position
  select(-Primary_Substance_Group) %>%
  rename(Primary_Substance_Group = substance_code)

# Merge data frames to replace age groups with age codes
admissions_data <- admissions_data %>%
  
  # Add column with county codes
  left_join(age_df, by = ("Age_Group")) %>%
  
  # Remove full names, rename substance codes, move columns to original position
  select(-Age_Group) %>%
  rename(Age_Group = age_code) %>%
  select(Year, County_of_Program_Location, Program_Category, 
         Service_Type, Age_Group, Primary_Substance_Group, Admissions)

# Check that the new data frame has the appropriate data and dimensions
head(admissions_data, 2)
dim(admissions_data)

```



### Question 4

Using the annualAdmissions() function developed in this question, I have displayed the total number of admissions in the state of New York from 2007-2021. This line chart displays a steady decline in admissions over time with a steep drop off in 2020 with admissions falling even lower 2021. While the gradual decline can be seen as positive progress regarding the frequency of chemical dependence in New Yorkers, the steep drop off in 2020 suggests that external factors may be influencing the data rather than a true decline of this magnitude. A likely candidate for this decline is COVID-19 restrictions that were implemented in 2020. New York famously battled early waves of the virus and as a result other medical treatments and programs, like chemical dependence treatment programs, faced limited capacities.

```{r message = FALSE, warning = FALSE}
# Function to determine number of admissions by year
annualAdmissions <- function(x) {
  admissions <- admissions_data %>%
    
    # Group by year and sum admissions in each year
    group_by(Year) %>%
    summarize(total_admissions = sum(Admissions))
  
  # Create a line chart showing the annual admissions over time
  ggplot(admissions, aes(x = Year, y = total_admissions)) +
    geom_line(color = "#1F4E78", alpha = 0.7) +
    geom_point(color = "#1F4E78", shape = 19) +
    labs(title = "Chemical Dependence Treatment Program Admissions",
         subtitle = "New York, 2007-2021",
         x = "Year",
         y = "Number of Admissions", 
         caption = "Display of total admissions over time shows the gradual decline in total admissions. 
         Source: https://data.ny.gov/api/views/ngbt-9rwf/rows.csv") +
    xlim(2007, 2021) +
    theme(text = element_text(family = "serif"),
          panel.grid.major.x = element_line(linewidth=0.1),
          panel.grid.major.y = element_line(linewidth = 0.3),
          panel.grid.minor = element_blank(),
          plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          plot.caption = element_text(size = 6),
          plot.margin = margin(10, 20, 10, 20, "pt"),
          plot.background = element_rect(linewidth = 1, color = "black"))
}

# Use function to create line chart and add annotations
annualAdmissions() +
  annotate(geom="point", 
           x = 2009, y = 311717, 
           size = 10, shape = 21, 
           fill="transparent",) +
  annotate(geom = "text", 
           x = 2011.8, y = 320000, 
           size = 3, family = "serif",
           label = "2009, Admissions peak = 311717") +
  scale_x_continuous(breaks = seq(2007, 2021, by = 2)) +
  scale_y_continuous(breaks = seq(175000, 325000, by = 25000),
                     limits = c(175000, 325000))

```


### Question 5

The chart below shows the percentage of total admissions throughout the state attributed to each county for each of the five counties with the highest percentages. It comes as no surprise that these counties are some of the most populous in the entire state, making up the bulk of New York City. In fact, each county corresponds with one of the five major burroughs of New York City, with Suffolk County being the only one not coextensive with a specific burrough. In sum, these 5 counties make up for 45.7% of all admissions.

```{r}
# Process original data frame
admissions_by_county <- admissions_data %>%
  
  # Rename County_of_Program_location to county_code for merge
  rename(county_code = County_of_Program_Location) %>%
  
  # Replace county_codes with full names for graphing label purposes
  left_join(county, by = ("county_code")) %>%
  
  # Group by county, sum admissions and calculate % of admissions by county
  group_by(County_of_Program_Location) %>%
  summarize(county_admissions = sum(Admissions)) %>%
  mutate(percentage_admissions = 100 * (county_admissions / sum(county_admissions))) %>%
  
  # Arrange % of admissions in descending order, slice top 5
  arrange(desc(percentage_admissions)) %>%
  slice_head(n = 5) 


# Define colors for bars, each county represented by color on county seal/flag
county_colors <- c("#ff6434", # Bronx
                   "#508cc4", # Kings (Brooklyn)
                   "#013585", # New York (Manhattan)
                   "#058200", # Queens
                   "#FFE716"  # Suffolk (Long Island)
                   )

# Create a bar chart showing the annual admissions over time
ggplot(admissions_by_county, aes(x = County_of_Program_Location, 
                                 y = percentage_admissions, 
                                 fill = County_of_Program_Location)) +
  geom_col(alpha = 0.6, color = "black", linewidth = 0.3) +
  labs(title = "Top 5 Counties by Percentage of Total Admissions",
       subtitle = "New York, 2007-2021",
       x = "County",
       y = "Percentage of Total Admissions", 
       caption = "This bar chart highlights how the top five counties in percentage of total admissions in the state account for nearly half of all admissions.
       Source: https://data.ny.gov/api/views/ngbt-9rwf/rows.csv") +
  scale_fill_manual(values = county_colors) +
  scale_y_continuous(expand = c(0.01, 0)) +
  theme(text = element_text(family = "serif"),
        legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linewidth = 0.2),
        panel.grid.minor.y = element_blank(),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(size = 6),
        plot.margin = margin(10, 20, 10, 20, "pt"),
        plot.background = element_rect(linewidth = 1, color = "black"))

# Provide summary statistic on percentage of total admissions in the state coming from the top 5 counties
admissions_by_county %>%
  summarize(pct_of_total_adm_from_top5 = sum(percentage_admissions))
```


### Question 6


The function grepl is used for pattern matching with regular expressions. It returns TRUE if a pattern is found in the specified string and FALSE otherwise. Here, it's being used to search for specific patterns within the Service_Type column of admissions_data.

```{r}
# filtering the data
rehab_facilities <- admissions_data[grepl("Inpatient Rehabilitation|Outpatient Rehabilitation|Residential Rehab for Youth|Residential Rehabilitation|Stabilization Rehab Reintegration|Rehab and Reintegration", admissions_data$Service_Type), ]


```
The pattern "Inpatient Rehabilitation|Outpatient Rehabilitation|Residential Rehab for Youth|Residential Rehabilitation|Stabilization Rehab Reintegration|Rehab and Reintegration" is a regular expression. The | character acts as a logical OR, so grepl will return TRUE for any string in Service_Type that matches any of these patterns.

The summarise function is used to create a summary statistic for each group created by group_by. Here, it's being used to calculate the most common substance for each age group.


```{r}
#Identify the Most Prominent Substance in Each Age Group
prominent_substance <- rehab_facilities %>%
  group_by(Age_Group) %>%
  summarise(most_common_substance = names(which.max(tapply(Admissions, Primary_Substance_Group, sum))))
```
The above function ensure to identify the most common substance for each Age group. This is done through via the max function in the last line.

```{r}
# visualization via barplot
ggplot(prominent_substance, aes(x = Age_Group, y = most_common_substance, fill = Age_Group)) +
  geom_bar(stat = "identity", alpha = 0.6, color = "black", linewidth = 0.3) + 
  scale_fill_manual(values = c("#ff6434", "#508cc4", "#013585", "#058200", "#FFE716", "#9932CC")) + 
  labs(title = "Most Common drug used per age group",
       x = "Age Group",
       y = "Most Common Substance", 
       caption = "Visualize which substance is used most by age group.
       Source: https://data.ny.gov/api/views/ngbt-9rwf/rows.csv") +
   theme(text = element_text(family = "serif"),
        legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linewidth = 0.2),
        panel.grid.minor.y = element_blank(),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(size = 6),
        plot.margin = margin(10, 20, 10, 20, "pt"),
        plot.background = element_rect(linewidth = 1, color = "black"))
```

From the produced graph above, we can see that for minors, the most common substance is marijuana, while in younger adutls (1-34) heroin is the most common. For those above the age of 34, alcohol is seen as the most common substance use. These patterns can inform public health strategies and resource allocation for treatment programs. 
